---
title: "test exam solution"
output: html_document
date: "2026-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercise 1.5

```{r}
ma2 <- arima.sim(model=list(ma=c(.1)), n=1000, sd=0.1) 
ts.plot(ma2)
```

Exercise 2.1

```{r}
set.seed(300)
```

```{r}
load("./Data/series.Rdata")
ts <- ts(series)
plot(ts)
```

I would say that the TS is stationary, as it seems to have a constant mean and variance throughout all 100 time steps.

2.2 Check it using kpss

```{r}
library(tseries)
kpss.test(ts)
```

we see that the p-value is greater than 0.05, making it highly likely that the ts is stationary

2.3 we look at the acf

```{r}
acf(ts)
```

MA(3) seems to be right, as the series seems to have autocorrelation with the 3 previous values, as the first three lags in the acf function cross the threshold. The small spikes at 15, 21 and 30 are a little above threshold, but are not relevant for the MA funciton

2.4

```{r}
ma2 <- arima(series, order=c(0,0,2))
print(ma2)
```

our model has three parameters:\
theta1 = ma1 = 0,55\
theta2 = ma2 = 0,36\
c = intercept = 0,061

aic is 2892.11

2.5

```{r}
predictions <- forecast(ma2, h = 5)

plot(series)
lines(predictions$mean, col =2)
```

2.6

```{r}
format_ts_to_tabular <- function(ts_data, window_size, step_ahead = 1) {
  n <- length(ts_data)
  num_rows <- n - window_size - step_ahead + 1
  
  X <- matrix(NA, nrow = num_rows, ncol = window_size)
  Y <- numeric(num_rows)
  
  for (i in 1:num_rows) {
    X[i, ] <- ts_data[i:(i + window_size - 1)]
    Y[i] <- ts_data[i + window_size + step_ahead - 1]
  }
  
  df <- data.frame(X, Y = Y)
  colnames(df)[1:window_size] <- paste0("lag_", window_size:1)
  return(df)
}

```

```{r}
library(randomForest)
df <- data.frame(embed(series, 6))
rf_model <- randomForest(X1 ~ ., data = df)

rf_preds <- numeric(10)
window <- tail(series, 5)

for(i in 1:10) {
  # Predict the next single step
  # (We reverse the window so the most recent data point is first, matching our training data)
  rf_preds[i] <- predict(rf_model, newdata = setNames(data.frame(t(rev(window))), names(df)[-1]))
  
  # Slide the window: drop the oldest point, add the brand new prediction
  window <- c(window[-1], rf_preds[i]) 
}


ts.plot(c(series,rf_preds))
lines(length(series) + 1:10, rf_preds, col=2)
     


```

3.2 calculate the distance autotically and double check

```{r}
load("./Data/tsdatabase.RData")

print(tsdatabase[1,])
print(tsdatabase[2,])

series1 <- tsdatabase[1,]
series2 <- tsdatabase[2,]

cost <- matrix(0, nrow = 5, ncol = 5)
cost[,1] = 10000
cost[1,] = 10000
cost[1,1] = 0

#main loop: for each entry distance plus min(up, left, up+left) from cost matrix
for (i in 2:nrow(cost)) {        # Outer loop: Rows
  for (j in 2:ncol(cost)) {      # Inner loop: Columns/Entries
    x <- cost[i-1,j]
    y <- cost[i,j-1]
    z <- cost[i-1,j-1]
    cost[i,j] <- abs(series1[i-1] - series2[j-1]) + min(x,y,z)
  }
}
cost
```

3.3.

we first get the dtw distances, and do the clustering usinf pam

```{r}
data <- tsdatabase
d_dtw <- dist(data, method = "dtw", step.pattern = symmetric1)
```

```{r}
k <- 3   # adapt

clus_dtw <- pam(d_dtw, k = k, diss = TRUE)

groups_dtw_kmedioids <- clus_dtw$clustering

print(groups_dtw_kmedioids)

```

reuse the distances to do the hierarchical clustering

```{r}
#DTW
hc_dtw <- hclust(d_dtw,  method="complete")

plot(hc_dtw)

groups_hc_dtw <-cutree(hc_dtw, k=3)
```

```{r}
# tsfeatures() needs a LIST of numeric vectors, not a data.frame!
series_list <- lapply(seq_len(nrow(data)), function(i) as.numeric(data[i, ]))

feats <- tsfeatures(series_list)
head(feats)

# REMOVE features with NA values (common exam requirement!)
feats <- feats[, colSums(is.na(feats)) == 0]

# k-means (works now because features form a Euclidean space)
groups_feat_km <- kmeans(feats, centers = 3)$cluster

# Or hierarchical on feature vectors
d_feat <- dist(feats, method = "euclidean")
hc_feat <- hclust(d_feat, method = "complete")
groups_feat_hc <- cutree(hc_feat, k = 3)

print(groups_feat_km)
```

```{r}
print(groups_feat_km)
print(groups_dtw)
print(groups_hc_dtw)
print(ground_truth)


ground_truth <- c(2, 3, 1, 1, 2, 3, 3, 2, 3, 2, 1, 1)

par(mfrow=c(2,1), mar=c(2,2,2,2))



```

```{r}
# Custom F-measure function (given in Lab 5)
F <- function(clus, true) {
  tab1 <- 2 * table(clus, true)
  tab2 <- outer(rowSums(tab1), colSums(tab1), `+`)
  result <- 1 / length(true) * sum(colSums(tab1) * apply((tab1 / tab2), 2, max))
  return(result)
}

# Usage: compare cluster labels vs ground truth
F(groups_feat_km, ground_truth)
F(groups_dtw, ground_truth)
F(groups_hc_dtw, ground_truth)
# Higher is better (max = 1)
```
