---
title: "ASTD – R Cheat Sheet for Exam Preparation"
author: "Compiled from Labs 1, 2, 3 & 5"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

------------------------------------------------------------------------

# 1. R Basics

## 1.1 Variables & Assignment

```{r}
x <- 5                      # assign value
x = 5                       # also works (less common in R)
y <- "hello"                # character string
z <- TRUE                   # logical
```

## 1.2 Vectors

```{r}
v <- c(1, 0, 2, 3, 1, 2)          # create vector
length(v)                           # length of vector
v[1]                                # first element (1-indexed!)
v[2:4]                              # elements 2 through 4
v[-1]                               # all except first element
```

## 1.3 Sequences & Repetition

```{r}
seq(from=1977, to=2001, by=0.25)    # sequence with step
1:10                                 # shorthand for seq(1, 10, 1)
rep(0, 34)                           # repeat 0, 34 times
rep(c(1,2,3), length.out=20)        # repeat vector, trim/extend to length 20
```

## 1.4 Math Operations

```{r}
abs(-3)              # absolute value
sqrt(16)             # square root
sum(v)               # sum of vector
mean(v)              # mean
floor(7.8)           # round down → 7
ceiling(7.2)         # round up → 8
min(v); max(v)       # min / max
```

## 1.5 Data Frames & Matrices

```{r}
df <- data.frame(X = c(1,2,3), Y = c(4,5,6))   # create data.frame
df$X                                              # access column
df[1, ]                                           # first row
df[, 2]                                           # second column
nrow(df); ncol(df)                                # dimensions
head(df)                                          # first 6 rows
colnames(df)                                      # column names

m <- matrix(NA, nrow=5, ncol=3)                   # create matrix
m[2, 3] <- 99                                     # assign element
```

## 1.6 Apply Family

```{r}
sapply(list_of_things, function(x) sum(x))   # apply & simplify to vector
lapply(1:5, function(i) i^2)                 # apply & return list
tapply(values, groups, mean)                 # apply by group (used for seasonal means!)
apply(matrix, 1, sum)                        # apply to rows (2 = columns)
```

## 1.7 Useful Utilities

```{r}
set.seed(42)                    # set random seed for reproducibility
cat("RMSE:", 0.5, "\n")        # print formatted output
print(x)                        # print object
paste0("lag_", 1:5)             # concatenate strings (no separator)
paste("a", "b", sep="-")        # concatenate with separator → "a-b"
which.min(v)                    # index of minimum value
which.max(v)                    # index of maximum value
```

## 1.8 Reading Data

```{r}
# Tab/space-separated files
data <- read.table("file.txt")
data <- read.table("file.txt", header=TRUE, sep=" ")

# CSV files
data <- read.table("file.csv", sep=",", header=TRUE)

# Load .Rdata files
load("file.Rdata")     # objects are loaded into the environment directly

# Access columns
data$ColumnName
data[, 2]              # second column
data[, c("col1","col2")]
```

------------------------------------------------------------------------

# 2. Creating & Manipulating Time Series Objects

## 2.1 Creating a `ts` Object

```{r}
# From a vector or column
my_ts <- ts(data_vector, start=1977, frequency=4)   # quarterly data (start year)
my_ts <- ts(data_vector, start=2007, frequency=12)  # monthly data
my_ts <- ts(data_vector, start=1945, frequency=12)  # monthly data

# frequency values:
#   1  = annual
#   4  = quarterly (trimestral)
#   12 = monthly
#   52 = weekly
```

## 2.2 Useful ts Functions

```{r}
start(my_ts)          # start time
end(my_ts)            # end time
frequency(my_ts)      # frequency
time(my_ts)           # time index vector (useful for regression!)
cycle(my_ts)          # seasonal cycle index (1,2,3,4,1,2,3,4,... for quarterly)
window(my_ts, start=1980, end=1990)   # subset time series by time range
length(my_ts)         # number of observations
as.numeric(my_ts)     # convert ts to plain numeric vector
as.ts(vector)         # convert vector to ts
```

## 2.3 Type Conversions

```{r}
as.numeric(CBF[1,])   # data.frame row → numeric vector
as.ts(residuals)       # vector → ts object
ts(vector, start=1977, frequency=4)   # vector → ts with time info
```

------------------------------------------------------------------------

# 3. Plotting

## 3.1 Basic Plotting

```{r}
plot(my_ts)                                          # simple time series plot
plot(my_ts, main="Title", ylab="Y label", xlab="X")  # with labels
plot(x, y, type="l")                                 # line plot
ts.plot(my_ts)                                       # alternative ts plot

# type options: "l"=line, "p"=points, "b"=both, "h"=histogram-like
```

## 3.2 Adding to Plots

```{r}
lines(time_vector, values, col="red", lwd=2)  # add line to existing plot
abline(v=c(1977:2000), lty=2, col=3)          # vertical dashed lines
abline(h=0, col="gray")                       # horizontal line
legend("topright", legend=c("A","B"), col=c("black","red"), lty=1, lwd=c(1,2))
```

## 3.3 Multiple Plots

```{r}
par(mfrow=c(2,1))                # 2 rows, 1 col of plots
par(mfrow=c(2,2), mar=c(2,2,2,2))  # 2x2 grid, small margins
# After this, each plot() fills the next panel
```

------------------------------------------------------------------------

# 4. Exploratory Analysis & Stationarity (Lab 1)

## 4.1 Trend Estimation – Linear Regression (Least Squares)

```{r}
t <- time(my_ts)                             # get time index
# or manually: t <- seq(from=1977.25, to=2001, by=0.25)

# Linear trend
model_lin <- lm(my_ts ~ t)

# Polynomial trend (degree 2)
model_poly2 <- lm(my_ts ~ poly(t, 2))

# Polynomial trend (degree 3)
model_poly3 <- lm(my_ts ~ poly(t, 3))

# Access fitted values & residuals
model_lin$fitted.values       # or model_lin$fitted
model_lin$residuals

# Plot trend + residuals
par(mfrow=c(2,1))
plot(t, my_ts, type="l", main="Trend estimation")
lines(t, model_lin$fitted, col=3)
plot(as.ts(model_lin$residuals), main="Residuals")
```

## 4.2 Trend Estimation – Moving Averages

```{r}
library(forecast)

ma5 <- ma(my_ts, order=5)    # moving average window = 5
ma7 <- ma(my_ts, order=7)    # moving average window = 7
# Larger window → smoother trend, but lose more endpoints

# Residuals after removing MA trend
residuals <- ts(my_ts - ma5, start=1977, frequency=4)
plot(residuals)
```

## 4.3 Trend Removal – Differencing

```{r}
diff1 <- diff(my_ts)             # first difference: removes linear trend
diff2 <- diff(diff(my_ts))       # second difference: removes quadratic trend
# Can also: diff(my_ts, differences=2)

# Wrap result back as ts with correct time info
diff1_ts <- ts(diff1, start=1977, frequency=4)
```

## 4.4 Stationarity Testing – KPSS Test

```{r}
library(tseries)

kpss.test(my_ts)
# H0: series is STATIONARY
# p-value > 0.05 → cannot reject stationarity → series MAY BE stationary
# p-value < 0.05 → reject H0 → series is NOT stationary

kpss.test(model_lin$residuals)    # test residuals after detrending
kpss.test(na.omit(decomp$random)) # test random component (remove NAs first!)
```

## 4.5 Seasonality Estimation – Seasonal Means Method

```{r}
# Step 1: compute mean per season
seasonal_means <- tapply(my_ts, cycle(my_ts), mean)

# Step 2: standardize (center to zero)
standardized <- seasonal_means - mean(seasonal_means)

# Step 3: replicate to full series length
seasonal_component <- ts(rep(standardized, length.out=length(my_ts)),
                         start=1980, frequency=4)

# Step 4: residuals
residuals <- my_ts - seasonal_component
```

## 4.6 Classical Decomposition (Trend + Seasonal + Random)

```{r}
# Additive decomposition: Y = Trend + Seasonal + Random
decomp_add <- decompose(my_ts, type="additive")

# Multiplicative decomposition: Y = Trend × Seasonal × Random
decomp_mul <- decompose(my_ts, type="multiplicative")

plot(decomp_add)    # plots all 4 components

# Access individual components
decomp_add$trend
decomp_add$seasonal
decomp_add$random       # residuals

# Test stationarity of residuals (remove NA from edges!)
kpss.test(na.omit(decomp_add$random))
```

------------------------------------------------------------------------

# 5. ARIMA Modeling & Forecasting (Lab 2)

## 5.1 Simulating ARIMA Processes

```{r}
library(forecast)
set.seed(42)

# AR(2) with params 0.9, -0.5
ar2 <- arima.sim(model=list(ar=c(0.9, -0.5)), n=4000)

# MA(2) with params -0.7, 0.1
ma2 <- arima.sim(model=list(ma=c(-0.7, 0.1)), n=4000)

# MA(1) with param 0.1  (as in test exam)
ma1 <- arima.sim(model=list(ma=c(0.1)), n=1000)

# ARMA(2,2) with AR=(0.4, 0.1), MA=(0.5, 0.2)
arma22 <- arima.sim(list(order=c(2,0,2), ar=c(0.4,0.1), ma=c(0.5,0.2)), n=4000)

# ARIMA(1,1,1) with AR=0.4, MA=0.5
arima111 <- arima.sim(list(order=c(1,1,1), ar=c(0.4), ma=c(0.5)), n=4000)

ts.plot(ar2)   # quick plot
```

## 5.2 ACF & PACF – Choosing Model Order

```{r}
# ACF → use to choose MA order q
#   MA(q): ACF cuts off after lag q
acf(my_ts)

# PACF → use to choose AR order p
#   AR(p): PACF cuts off after lag p
pacf(my_ts)

# RULE: values outside the blue dashed confidence bands are "significant"
# Choose the SMALLEST lag after which ALL values are inside the bands
```

## 5.3 Fitting ARIMA Models

```{r}
# Manually specify order
arima(my_ts, order=c(2, 0, 0))   # AR(2)
arima(my_ts, order=c(0, 0, 2))   # MA(2)
arima(my_ts, order=c(1, 1, 1))   # ARIMA(1,1,1)

# Automatic model selection (uses AICc by default)
model <- auto.arima(my_ts, seasonal=FALSE)

# Restrict search space
auto.arima(my_ts, max.p=5, max.q=0, d=0, seasonal=FALSE)  # AR only
auto.arima(my_ts, max.p=0, max.q=5, d=0, seasonal=FALSE)  # MA only
auto.arima(my_ts, max.p=5, max.q=5, d=0, seasonal=FALSE)  # ARMA
auto.arima(my_ts, max.p=5, max.q=5, seasonal=FALSE)        # ARIMA (auto d)
```

## 5.4 Residual Analysis (Model Validation)

```{r}
# Combined check: ACF, histogram, Ljung-Box test
checkresiduals(model)
# Goal: residuals should look like WHITE NOISE
#   → ACF values inside confidence bands (no autocorrelation)
#   → Ljung-Box p-value > 0.05 (cannot reject independence)
#   → Histogram roughly bell-shaped (normal)

# Normality test
shapiro.test(model$residuals)
# H0: data is normally distributed
# p-value > 0.05 → cannot reject normality → residuals MAY BE normal
```

## 5.5 Forecasting

```{r}
predictions <- forecast(model, h=5)    # predict next 5 time points
predictions                             # shows point forecasts + confidence intervals
plot(predictions)                       # plot with fan chart

# Access components
predictions$mean      # point forecasts
predictions$lower     # lower CI bounds
predictions$upper     # upper CI bounds
```

------------------------------------------------------------------------

# 6. ML Regression for Time Series (Lab 3)

## 6.1 Sliding Window – Transform TS to Tabular Data

```{r}
# PURPOSE: classical ML algorithms need tabular (X, Y) data
# A sliding window creates lag features from the time series

format_ts_to_tabular <- function(ts_data, window_size, step_ahead = 1) {
  n <- length(ts_data)
  num_rows <- n - window_size - step_ahead + 1
  
  X <- matrix(NA, nrow = num_rows, ncol = window_size)
  Y <- numeric(num_rows)
  
  for (i in 1:num_rows) {
    X[i, ] <- ts_data[i:(i + window_size - 1)]       # features = past values
    Y[i]   <- ts_data[i + window_size + step_ahead - 1]  # target = future value
  }
  
  df <- data.frame(X, Y = Y)
  colnames(df)[1:window_size] <- paste0("lag_", window_size:1)
  return(df)
}

# Usage
w <- 12   # 12 months = 1-year window
tabular_data <- format_ts_to_tabular(my_ts, window_size = w)
# For 3-step ahead:
tabular_data_3 <- format_ts_to_tabular(my_ts, window_size = w, step_ahead = 3)
```

## 6.2 Train/Test Split (Temporal Order!)

```{r}
# IMPORTANT: for time series, split preserving temporal order (no random shuffle!)
n_train <- floor(0.7 * nrow(tabular_data))
train_data <- tabular_data[1:n_train, ]
test_data  <- tabular_data[(n_train + 1):nrow(tabular_data), ]
```

## 6.3 Time Series Cross-Validation with caret

```{r}
library(caret)

# createTimeSlices: respects chronological order
time_slices <- createTimeSlices(y = train_data$Y,
                                initialWindow = 36,    # min training window
                                horizon = 12,          # prediction horizon in CV
                                fixedWindow = FALSE)   # growing window

train_control <- trainControl(method = "timeslice",
                               index = time_slices$train,
                               indexOut = time_slices$test)
```

## 6.4 Training Models with caret

```{r}
library(caret)
set.seed(42)

# Linear Regression
lm_model <- train(Y ~ ., data = train_data,
                  method = "lm",
                  trControl = train_control)

# k-Nearest Neighbors (with hyperparameter search)
knn_model <- train(Y ~ ., data = train_data,
                   method = "knn",
                   trControl = train_control,
                   tuneGrid = expand.grid(k = 1:10))

# Other methods: "svmRadial", "rf" (random forest), "ridge", "lasso", etc.
```

## 6.5 Prediction & Evaluation (RMSE)

```{r}
# Predict on test set
lm_preds  <- predict(lm_model, newdata = test_data)
knn_preds <- predict(knn_model, newdata = test_data)

# RMSE (Root Mean Squared Error)
rmse_lm  <- RMSE(lm_preds, test_data$Y)
rmse_knn <- RMSE(knn_preds, test_data$Y)

cat("RMSE LM:", rmse_lm, "\n")
cat("RMSE KNN:", rmse_knn, "\n")

# Pick best
best_preds <- if (rmse_lm < rmse_knn) lm_preds else knn_preds
```

## 6.6 Plotting Predictions vs Actual

```{r}
# Reconstruct test time indices
original_time <- time(my_ts)
test_time <- original_time[(n_train + w + 1):length(original_time)]

plot(my_ts, type="l", main="Forecasting", ylab="Value", xlab="Time")
lines(test_time, best_preds, col="red", lwd=2)
legend("topright", legend=c("True","Predicted"), col=c("black","red"), lty=1)
```

------------------------------------------------------------------------

# 7. Distances for Time Series (Lab 5)

## 7.1 Euclidean & Manhattan Distance

```{r}
library(proxy)

# Between two time series (store as rows of a data.frame)
data <- rbind(series1, series2)   # or: data <- df[c(1,2), ]

dist(data, method="euclidean")
dist(data, method="manhattan")

# Manual calculation
euclidean_dist <- sqrt(sum((Xt - Yt)^2))
manhattan_dist <- sum(abs(Xt - Yt))
```

## 7.2 DTW Distance (Dynamic Time Warping)

```{r}
library(dtw)

# Using dist() from proxy
dist(data, method="dtw", step.pattern=symmetric1)

# Using dtw() directly (gives more info)
alignment <- dtw(x = as.numeric(series1),
                 y = as.numeric(series2),
                 keep.internals = TRUE)

alignment$distance         # the DTW distance value

# Visualization
plot(alignment, type="twoway", off=2, match.lty=2, match.indices=20)
plot(alignment, type="threeway")
```

## 7.3 DTW with Sakoe-Chiba Band

```{r}
alignment_banded <- dtw(x = as.numeric(series1),
                        y = as.numeric(series2),
                        step.pattern = symmetric1,
                        window.type = "sakoechiba",
                        window.size = 10,
                        keep.internals = TRUE)

plot(alignment_banded, type="density")   # shows allowed region
```

## 7.4 Count DTW Paths

```{r}
countPaths(alignment)    # number of optimal alignment paths
```

## 7.5 DTW Cost Matrix by Hand

```{r}
# Manual DTW cost matrix computation
Xt <- c(1, 0, 2, 3, 1, 2)
Yt <- c(0, 1, 2, 1, 0, 3)

n <- length(Xt) + 1
m <- length(Yt) + 1
cost <- matrix(0, nrow = n, ncol = m)
cost[, 1] <- Inf     # first column = infinity (except [1,1])
cost[1, ] <- Inf     # first row = infinity (except [1,1])
cost[1, 1] <- 0      # starting point

for (i in 2:n) {
  for (j in 2:m) {
    cost[i, j] <- abs(Xt[i-1] - Yt[j-1]) + min(cost[i-1, j],
                                                  cost[i, j-1],
                                                  cost[i-1, j-1])
  }
}
cost[n, m]   # final DTW distance
```

## 7.6 Pairwise Distance Matrices

```{r}
# All pairwise distances within one dataset
d_euc <- dist(CBF, method="euclidean")
d_dtw <- dist(CBF, method="dtw", step.pattern=symmetric1)

# Pairwise distances between two matrices A and B
distances <- matrix(NA, nrow=nrow(A), ncol=nrow(B))
for (i in 1:nrow(A)) {
  for (j in 1:nrow(B)) {
    distances[i,j] <- dist(rbind(A[i,], B[j,]), method="euclidean")
  }
}
```

------------------------------------------------------------------------

# 8. Clustering Time Series (Lab 5)

## 8.1 K-Medoids (PAM) – Works with Any Distance

```{r}
library(cluster)

# Input: precomputed distance matrix, diss=TRUE
clus <- pam(d_euc, k=3, diss=TRUE)

clus$clustering     # cluster assignments (vector of integers)
clus$medoids        # medoid indices

# With DTW distance
clus_dtw <- pam(d_dtw, k=3, diss=TRUE)$clustering
```

## 8.2 Hierarchical Clustering

```{r}
# Input: distance matrix
hc <- hclust(d_euc, method="complete")
# method: "complete", "single", "average", "ward.D2"

plot(hc)   # dendrogram → visually choose k

# Cut dendrogram to get k clusters
groups <- cutree(hc, k=3)
```

## 8.3 K-Means – Only with Euclidean / Feature-Based

```{r}
# Direct k-means (only works with raw features, Euclidean distance)
result <- kmeans(feature_matrix, centers=3)
result$cluster     # cluster assignments
```

## 8.4 Feature-Based Clustering with tsfeatures

```{r}
library(tsfeatures)

# Convert data.frame rows to list of numeric vectors
ts_list <- lapply(seq_len(nrow(df)), function(i) as.numeric(df[i,]))

# Extract features
features <- tsfeatures(ts_list)
head(features)

# Handle NAs
features[is.na(features)] <- 0

# Now cluster using k-means (features are tabular!)
clusters <- kmeans(features, centers=3)$cluster
```

------------------------------------------------------------------------

# 9. Evaluation

## 9.1 F-Measure for Clustering (External Evaluation)

```{r}
# Custom F-measure function (given in Lab 5)
F <- function(clus, true) {
  tab1 <- 2 * table(clus, true)
  tab2 <- outer(rowSums(tab1), colSums(tab1), `+`)
  result <- 1 / length(true) * sum(colSums(tab1) * apply((tab1 / tab2), 2, max))
  return(result)
}

# Usage: compare cluster labels vs ground truth
F(cluster_result, true_labels)
# Higher is better (max = 1)
```

## 9.2 RMSE (Regression / Forecasting)

```{r}
# Using caret
library(caret)
RMSE(predicted, actual)

# Manual
rmse <- sqrt(mean((predicted - actual)^2))
```

## 9.3 Stratified Sampling (caret)

```{r}
library(caret)

# Stratified partitioning (preserves class proportions)
indices <- createDataPartition(data$class_column, p=0.2, list=FALSE)
subset_A <- data[indices, ]

# Simple random sampling (does NOT preserve proportions)
random_idx <- sample(1:nrow(data), size=0.2*nrow(data))
```

------------------------------------------------------------------------

# 10. Quick Reference – Package Cheat Sheet

| Package | Key Functions | Purpose |
|----|----|----|
| `base` | `ts()`, `diff()`, `plot()`, `lm()`, `acf()`, `pacf()` | Core R time series & regression |
| `forecast` | `ma()`, `auto.arima()`, `forecast()`, `checkresiduals()` | ARIMA modeling & forecasting |
| `tseries` | `kpss.test()` | Stationarity test |
| `caret` | `train()`, `predict()`, `RMSE()`, `trainControl()` | ML training & evaluation |
| `caret` | `createTimeSlices()`, `createDataPartition()` | Temporal CV & stratified splits |
| `proxy` | `dist()` (extended methods) | Distance computation |
| `dtw` | `dtw()`, `countPaths()` | DTW distance & alignment |
| `cluster` | `pam()` | K-medoids clustering |
| `stats` | `hclust()`, `cutree()`, `kmeans()`, `decompose()`, `shapiro.test()` | Clustering & decomposition |
| `tsfeatures` | `tsfeatures()` | Feature extraction from TS |

------------------------------------------------------------------------

# 11. Common Exam Patterns

## Pattern A: "Load data → create ts → explore → decompose"

```{r}
# 1. Load
data <- read.table("file.txt")
my_ts <- ts(data[, 2], start=1977, frequency=4)

# 2. Plot
plot(my_ts, main="Title")

# 3. Estimate trend (choose one)
t <- time(my_ts)
model <- lm(my_ts ~ poly(t, 2))            # polynomial
trend_ma <- ma(my_ts, order=5)              # moving average
differenced <- diff(my_ts)                  # differencing

# 4. Check stationarity of residuals
library(tseries)
kpss.test(model$residuals)

# 5. Full decomposition
decomp <- decompose(my_ts, type="additive")
plot(decomp)
kpss.test(na.omit(decomp$random))
```

## Pattern B: "Fit ARIMA → check residuals → forecast"

```{r}
library(forecast)

# 1. Check stationarity, difference if needed
diff_ts <- diff(my_ts)

# 2. ACF/PACF to decide order (or use auto.arima)
acf(diff_ts)
pacf(diff_ts)

# 3. Fit model
model <- auto.arima(my_ts, seasonal=FALSE)

# 4. Check residuals
checkresiduals(model)
shapiro.test(model$residuals)

# 5. Forecast
preds <- forecast(model, h=10)
plot(preds)
```

## Pattern C: "Simulate → plot → identify"

```{r}
# Simulate a process
sim <- arima.sim(model=list(ma=c(0.1)), n=1000)     # MA(1)
sim <- arima.sim(model=list(ar=c(0.9, -0.5)), n=1000)  # AR(2)
sim <- arima.sim(list(order=c(1,1,1), ar=0.4, ma=0.5), n=1000)  # ARIMA(1,1,1)

ts.plot(sim)
acf(sim)
pacf(sim)
```

## Pattern D: "Sliding window → train ML → evaluate RMSE"

```{r}
library(caret)

# 1. Transform to tabular
tabular <- format_ts_to_tabular(my_ts, window_size=12)

# 2. Temporal split
n <- floor(0.7 * nrow(tabular))
train <- tabular[1:n, ]
test  <- tabular[(n+1):nrow(tabular), ]

# 3. Train
model <- train(Y ~ ., data=train, method="knn",
               trControl=trainControl(method="timeslice",
                                       index=createTimeSlices(train$Y, 36, 12)$train,
                                       indexOut=createTimeSlices(train$Y, 36, 12)$test))

# 4. Evaluate
preds <- predict(model, newdata=test)
RMSE(preds, test$Y)
```

## Pattern E: "Compute distances → cluster → evaluate"

```{r}
library(proxy); library(dtw); library(cluster)

# 1. Distance matrix
d <- dist(data_matrix, method="dtw", step.pattern=symmetric1)

# 2. Cluster
groups <- pam(d, k=3, diss=TRUE)$clustering

# 3. Evaluate (if labels available)
F(groups, true_labels)

# Or hierarchical:
hc <- hclust(d, method="complete")
plot(hc)
groups_hc <- cutree(hc, k=3)
```

------------------------------------------------------------------------

# 12. Key Statistical Tests Summary

| Test | Function | H₀ (Null Hypothesis) | Reject when |
|----|----|----|----|
| **KPSS** | `kpss.test(x)` | Series is **stationary** | p-value \< 0.05 |
| **Ljung-Box** | `checkresiduals(m)` | Residuals are **independent** | p-value \< 0.05 |
| **Shapiro-Wilk** | `shapiro.test(x)` | Data is **normally distributed** | p-value \< 0.05 |

> **Remember**: In all tests, a **high p-value** means you **cannot reject** H₀ — it does NOT prove H₀ is true.

------------------------------------------------------------------------

# 13. Gotchas & Tips

1.  **`na.omit()`** – Always use before `kpss.test()` on `decompose()$random` (has NAs at edges)
2.  **`diss=TRUE`** – Required in `pam()` when passing a precomputed distance matrix
3.  **`step.pattern=symmetric1`** – Always specify for DTW to match the method from class
4.  **Temporal train/test split** – Never shuffle time series data! Always keep chronological order
5.  **`set.seed()`** – Use before any randomized operation for reproducibility
6.  **`cycle()`** returns the seasonal period index — use with `tapply()` for seasonal means
7.  **`diff()` shortcut** – `diff(x, differences=2)` is equivalent to `diff(diff(x))`
8.  **Feature clustering** – After `tsfeatures()`, the data is tabular → can use `kmeans()` directly
