---
title: Time series forecasting with classical Machine Learning algorithms
author: Usue Mori
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H',fig.align="center")
library(knitr)
hook_output = knit_hooks$get('output')

knit_hooks$set(document = function(x) {
  sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)
})


# knit_theme$set("earendel")
# knit_theme$set("print")
 knitr::opts_chunk$set(linewidth=90, comment=NA,background='grey92') 
 

```

<br>

In this tutorial we will learn to perform forecasts on time series using classical regression algorithms from machine learning. As an example, we are going to use a time series that represents the long-term interest rates in E.E.U.U from January 2007 to March 2018. [^1]

[^1]: Time series extracted from <http://rwanjohi.rbind.io/2018/04/05/time-series-forecasting-using-lstm-in-r/>).

```{r, warning=FALSE}
interestsEEUU <- read.table("./Data/MEI_FIN_23092019161922426.csv", sep=",", header=TRUE)
interestsEEUU <- ts(interestsEEUU[interestsEEUU$Country=="United States" 
                    & interestsEEUU$Tipo=="IRLT", c("Value")], 
                    start=2007, frequency=12)
plot(interestsEEUU)
```

# Exercises

-   In order to evaluate our models in a honest way, we will use a hold-out method, but we will respect the temporal order to the data. This means that we will use the first 70% data points (first 8 years) of the time series to train the models are the remaining 30% to test them.

NOTE: The `caret` package proposes specific methods to generate train/test splits in time series. It also proposes ways to perform cross-validation with temporal data. For more information go to \url{https://otexts.com/fpp2/accuracy.html}.

-   As we have seen in the theory, classical regression algorithms are not designed to perform predictions on time series, they are designed for tabular data. So, in order to apply them to temporal data, we must transform our time series into a tabular form, using a sliding window. Program a function in R that will format the data in R so that it is suitable for classical regression algorithms. This function will take two parameters as input: the time series and the size of the sliding window.

```{r, warning=FALSE}
library(caret)
# 1. Transform time series to tabular form
# This function dynamically creates feature columns (lags) and a target variable
format_ts_to_tabular <- function(ts_data, window_size, step_ahead = 1) {
  n <- length(ts_data)
  num_rows <- n - window_size - step_ahead + 1
  
  X <- matrix(NA, nrow = num_rows, ncol = window_size)
  Y <- numeric(num_rows)
  
  for (i in 1:num_rows) {
    X[i, ] <- ts_data[i:(i + window_size - 1)]
    Y[i] <- ts_data[i + window_size + step_ahead - 1]
  }
  
  df <- data.frame(X, Y = Y)
  colnames(df)[1:window_size] <- paste0("lag_", window_size:1)
  return(df)
}

# Apply using a 1-year sliding window (w = 12 months)
w <- 12
tabular_data <- format_ts_to_tabular(interestsEEUU, window_size = w)

# 2. Train/Test Split (70% train, 30% test) matching temporal order
n_train <- floor(0.7 * nrow(tabular_data))
train_data <- tabular_data[1:n_train, ]
test_data <- tabular_data[(n_train + 1):nrow(tabular_data), ]

# Prepare cross-validation setup for time series (caret)
# We use createTimeSlices to respect chronological order natively
time_slices <- createTimeSlices(y = train_data$Y, 
                                initialWindow = 36, # Start with 3 years basic training
                                horizon = 12,       # Predict 1 year forward in CV
                                fixedWindow = FALSE)

train_control <- trainControl(method = "timeslice",
                              index = time_slices$train,
                              indexOut = time_slices$test)
```

-   Apply different regression algorithms from the `caret` package to the formatted time series (use only the training set for fitting the models).

```{r train_models, warning=FALSE, message=FALSE}
# Train a Linear Regression model
set.seed(42)
lm_model <- train(Y ~ ., data = train_data, 
                  method = "lm", 
                  trControl = train_control)

# Train a k-Nearest Neighbors model
set.seed(42)
knn_model <- train(Y ~ ., data = train_data, 
                   method = "knn", 
                   trControl = train_control,
                   tuneGrid = expand.grid(k = 1:10)) # Search best k
```

-   Evaluate the models by using the test set and computing the RMSE and represent the predictions of the best model together with the time series.

```{r evaluate_models, warning=FALSE, message=FALSE}
# Predictions on test set
lm_preds <- predict(lm_model, newdata = test_data)
knn_preds <- predict(knn_model, newdata = test_data)

# Compute Root Mean Squared Error
rmse_lm <- RMSE(lm_preds, test_data$Y)
rmse_knn <- RMSE(knn_preds, test_data$Y)

cat("RMSE for Linear Regression:", rmse_lm, "\n")
cat("RMSE for k-Nearest Neighbors:", rmse_knn, "\n")

# Identify the best performance based on RMSE
best_predicts <- if (rmse_lm < rmse_knn) lm_preds else knn_preds
best_model_name <- if (rmse_lm < rmse_knn) "Linear Regression" else "kNN"

# Plotting the series with predictions
# Reconstruct the correct time indices for test set predictions
original_time <- time(interestsEEUU)
# The test set starts after `n_train` tabular rows, and shifts by `w` lags + 1 step ahead
test_time <- original_time[(n_train + w + 1):length(original_time)]

plot(interestsEEUU, type = "l", 
     main = paste("1-Step Ahead Forecasting (", best_model_name, ")", sep=""), 
     ylab = "Interest Rate", xlab = "Time", col = "black")
lines(test_time, best_predicts, col = "red", lwd = 2)
legend("topright", legend = c("True Value", "Predictions"), 
       col = c("black", "red"), lty = 1, lwd = c(1, 2))
```

# Exam prep

-   Repeat the previous exercise to perform 3-step ahead predictions.

```{r extra_work, warning=FALSE, message=FALSE}
# 1. Format data for 3-steps ahead predicting
step <- 3
data_3step <- format_ts_to_tabular(interestsEEUU, window_size = w, step_ahead = step)

# 2. Split into train & test
n_train_3 <- floor(0.7 * nrow(data_3step))
train_data_3 <- data_3step[1:n_train_3, ]
test_data_3 <- data_3step[(n_train_3 + 1):nrow(data_3step), ]

# Prepare time slices matching the split length
slices_3 <- createTimeSlices(y = train_data_3$Y, 
                             initialWindow = 36, horizon = 12, fixedWindow = FALSE)
train_ctrl_3 <- trainControl(method = "timeslice",
                             index = slices_3$train, indexOut = slices_3$test)

# 3. Train models
set.seed(42)
lm_model_3 <- train(Y ~ ., data = train_data_3, method = "lm", trControl = train_ctrl_3)
set.seed(42)
knn_model_3 <- train(Y ~ ., data = train_data_3, method = "knn", trControl = train_ctrl_3)

# 4. Evaluate Models
lm_preds_3 <- predict(lm_model_3, newdata = test_data_3)
knn_preds_3 <- predict(knn_model_3, newdata = test_data_3)

rmse_lm_3 <- RMSE(lm_preds_3, test_data_3$Y)
rmse_knn_3 <- RMSE(knn_preds_3, test_data_3$Y)

cat("3-Step Ahead RMSE for Linear Regression:", rmse_lm_3, "\n")
cat("3-Step Ahead RMSE for k-Nearest Neighbors:", rmse_knn_3, "\n")

best_predicts_3 <- if (rmse_lm_3 < rmse_knn_3) lm_preds_3 else knn_preds_3
best_model_name_3 <- if (rmse_lm_3 < rmse_knn_3) "Linear Regression" else "kNN"

# 5. Visualizer
test_time_3 <- original_time[(n_train_3 + w + step):length(original_time)]

plot(interestsEEUU, type = "l", 
     main = paste("3-Step Ahead Forecasting (", best_model_name_3, ")", sep=""), 
     ylab = "Interest Rate", xlab = "Time", col = "black")
lines(test_time_3, best_predicts_3, col = "red", lwd = 2)
legend("topright", legend = c("True Value", "Predictions"), 
       col = c("black", "red"), lty = 1, lwd = c(1, 2))
```

---
title: "test exam solution"
output: html_document
date: "2026-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercise 1.5

```{r}
ma2 <- arima.sim(model=list(ma=c(.1)), n=1000, sd=0.1) 
ts.plot(ma2)
```

Exercise 2.1

```{r}
set.seed(300)
```

```{r}
load("./Data/series.Rdata")
ts <- ts(series)
plot(ts)
```

I would say that the TS is stationary, as it seems to have a constant mean and variance throughout all 100 time steps.

2.2 Check it using kpss

```{r}
library(tseries)
kpss.test(ts)
```

we see that the p-value is greater than 0.05, making it highly likely that the ts is stationary

2.3 we look at the acf

```{r}
acf(ts)
```

MA(3) seems to be right, as the series seems to have autocorrelation with the 3 previous values, as the first three lags in the acf function cross the threshold. The small spikes at 15, 21 and 30 are a little above threshold, but are not relevant for the MA funciton

2.4

```{r}
ma2 <- arima(series, order=c(0,0,2))
print(ma2)
```

our model has three parameters:\
theta1 = ma1 = 0,55\
theta2 = ma2 = 0,36\
c = intercept = 0,061

aic is 2892.11

2.5

```{r}
predictions <- forecast(ma2, h = 5)

plot(series)
lines(predictions$mean, col =2)
```

2.6

```{r}
format_ts_to_tabular <- function(ts_data, window_size, step_ahead = 1) {
  n <- length(ts_data)
  num_rows <- n - window_size - step_ahead + 1
  
  X <- matrix(NA, nrow = num_rows, ncol = window_size)
  Y <- numeric(num_rows)
  
  for (i in 1:num_rows) {
    X[i, ] <- ts_data[i:(i + window_size - 1)]
    Y[i] <- ts_data[i + window_size + step_ahead - 1]
  }
  
  df <- data.frame(X, Y = Y)
  colnames(df)[1:window_size] <- paste0("lag_", window_size:1)
  return(df)
}

```

```{r}
library(randomForest)
df <- data.frame(embed(series, 6))
rf_model <- randomForest(X1 ~ ., data = df)

rf_preds <- numeric(10)
window <- tail(series, 5)

for(i in 1:10) {
  # Predict the next single step
  # (We reverse the window so the most recent data point is first, matching our training data)
  rf_preds[i] <- predict(rf_model, newdata = setNames(data.frame(t(rev(window))), names(df)[-1]))
  
  # Slide the window: drop the oldest point, add the brand new prediction
  window <- c(window[-1], rf_preds[i]) 
}


ts.plot(c(series,rf_preds))
lines(length(series) + 1:10, rf_preds, col=2)
     


```

3.2 calculate the distance autotically and double check

```{r}
load("./Data/tsdatabase.RData")

print(tsdatabase[1,])
print(tsdatabase[2,])

series1 <- tsdatabase[1,]
series2 <- tsdatabase[2,]

cost <- matrix(0, nrow = 5, ncol = 5)
cost[,1] = 10000
cost[1,] = 10000
cost[1,1] = 0

#main loop: for each entry distance plus min(up, left, up+left) from cost matrix
for (i in 2:nrow(cost)) {        # Outer loop: Rows
  for (j in 2:ncol(cost)) {      # Inner loop: Columns/Entries
    x <- cost[i-1,j]
    y <- cost[i,j-1]
    z <- cost[i-1,j-1]
    cost[i,j] <- abs(series1[i-1] - series2[j-1]) + min(x,y,z)
  }
}
cost
```

3.3.

we first get the dtw distances, and do the clustering usinf pam

```{r}
data <- tsdatabase
d_dtw <- dist(data, method = "dtw", step.pattern = symmetric1)
```

```{r}
k <- 3   # adapt

clus_dtw <- pam(d_dtw, k = k, diss = TRUE)

groups_dtw_kmedioids <- clus_dtw$clustering

print(groups_dtw_kmedioids)

```

reuse the distances to do the hierarchical clustering

```{r}
#DTW
hc_dtw <- hclust(d_dtw,  method="complete")

plot(hc_dtw)

groups_hc_dtw <-cutree(hc_dtw, k=3)
```

```{r}
# tsfeatures() needs a LIST of numeric vectors, not a data.frame!
series_list <- lapply(seq_len(nrow(data)), function(i) as.numeric(data[i, ]))

feats <- tsfeatures(series_list)
head(feats)

# REMOVE features with NA values (common exam requirement!)
feats <- feats[, colSums(is.na(feats)) == 0]

# k-means (works now because features form a Euclidean space)
groups_feat_km <- kmeans(feats, centers = 3)$cluster

# Or hierarchical on feature vectors
d_feat <- dist(feats, method = "euclidean")
hc_feat <- hclust(d_feat, method = "complete")
groups_feat_hc <- cutree(hc_feat, k = 3)

print(groups_feat_km)
```

```{r}
print(groups_feat_km)
print(groups_dtw)
print(groups_hc_dtw)
print(ground_truth)


ground_truth <- c(2, 3, 1, 1, 2, 3, 3, 2, 3, 2, 1, 1)

par(mfrow=c(2,1), mar=c(2,2,2,2))



```

```{r}
# Custom F-measure function (given in Lab 5)
F <- function(clus, true) {
  tab1 <- 2 * table(clus, true)
  tab2 <- outer(rowSums(tab1), colSums(tab1), `+`)
  result <- 1 / length(true) * sum(colSums(tab1) * apply((tab1 / tab2), 2, max))
  return(result)
}

# Usage: compare cluster labels vs ground truth
F(groups_feat_km, ground_truth)
F(groups_dtw, ground_truth)
F(groups_hc_dtw, ground_truth)
# Higher is better (max = 1)
```

# References

-   Kuhn, M. (2008). Caret package. Journal of Statistical Software, 28(5)
